Appendix: Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-26 appendix.qmd
Clear the repo

```{r}
#| warning: false
#| echo: false
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)

library(Rtsne)
library(umap)
library(phateR)
library(reticulate)
library(patchwork)

library(grid)

# install.packages("Seurat")
# remotes::install_github("satijalab/seurat-data")
#library(SeuratData) ## For the application

# use_python("~/miniforge3/envs/pcamp_env/bin/python")
# use_condaenv("pcamp_env")
# 
# reticulate::source_python(paste0(here::here(), "/examples_for_the_paper/function_scripts/Fit_PacMAP_code.py"))
# reticulate::source_python(paste0(here::here(), "/examples_for_the_paper/function_scripts/Fit_TriMAP_code.py"))

set.seed(20230531)

source("quollr_code.R", local = TRUE)
source("nldr_code.R", local = TRUE)
```

@@

data <- read_rds("data/s_curve/s_curve.rds")
```

@@

training_data <- read_rds("data/s_curve/s_curve_training.rds")
test_data <- read_rds("data/s_curve/s_curve_test.rds")
```

```{r}
#| echo: false

UMAP_data <- read_rds("data/s_curve/s_curve_umap.rds")

num_bins_x <- calculate_effective_x_bins(.data = UMAP_data, x = UMAP1,
                           cell_area = 1)

shape_value <- calculate_effective_shape_value(.data = UMAP_data, 
                                             x = UMAP1, y = UMAP2)

## To extract bin centroids
hexbin_data_object <-extract_hexbin_centroids(nldr_df = UMAP_data, num_bins = num_bins_x, shape_val = shape_value)

df_bin_centroids <- hexbin_data_object$hexdf_data

##########

## Data set with all possible centroids in the hexagonal grid

full_centroid_df <- generate_full_grid_centroids(df_bin_centroids)

## To map hexID to hexbin centroids in the full grid

vec1 <- stats::setNames(rep("", 2), c("x", "y"))  ## Define column names

full_grid_with_hexbin_id <- dplyr::bind_rows(vec1)[0, ]
full_grid_with_hexbin_id <- full_grid_with_hexbin_id |>
  dplyr::mutate_if(is.character, as.numeric)

for(i in 1:length(sort(unique(full_centroid_df$y)))){

  ## Filter the data set with specific y value
  specific_y_val_df <- full_centroid_df |>
    dplyr::filter(y == sort(unique(full_centroid_df$y))[i])

  ordered_x_df <- specific_y_val_df |>
    dplyr::arrange(x)

  full_grid_with_hexbin_id <- dplyr::bind_rows(full_grid_with_hexbin_id, ordered_x_df)

@@

full_grid_with_hexbin_id <- full_grid_with_hexbin_id |>
  dplyr::mutate(hexID = row_number())

full_grid_with_hexbin_id <- full_grid_with_hexbin_id |>
  dplyr::rename("c_x" = "x",
         "c_y" = "y")

full_grid_with_hexbin_id <- dplyr::full_join(full_grid_with_hexbin_id, df_bin_centroids, by = c("hexID" = "hexID")) |>
  dplyr::select(-c(x, y)) #|> 
  #dplyr::mutate(counts = tidyr::replace_na(counts, 0))

full_grid_with_hexbin_id <- full_grid_with_hexbin_id |>
    dplyr::mutate(std_counts = counts/max(counts, na.rm = TRUE))

## Generate all coordinates of hexagons
hex_grid <- full_hex_grid(full_centroid_df)

full_grid_with_polygon_id_df <- map_polygon_id(full_grid_with_hexbin_id, hex_grid)

full_grid_with_hexbin_id_rep <- full_grid_with_polygon_id_df |>
  dplyr::slice(rep(1:n(), each = 6)) |>
  dplyr::arrange(polygon_id)

hex_full_count_df <- dplyr::bind_cols(hex_grid, full_grid_with_hexbin_id_rep)

##########

min_std_cell_threshold <- 0.05

df_bin_centroids_all <- hexbin_data_object$hexdf_data ## All the centroids without removing low-density hexagons

df_bin_centroids <- df_bin_centroids |>
  dplyr::mutate(stand_cell_count = counts/max(counts)) |>
  dplyr::filter(stand_cell_count > min_std_cell_threshold)

UMAP_data_with_hb_id <- UMAP_data |> 
  dplyr::mutate(hb_id = hexbin_data_object$hb_data@cID)

## To generate a data set with high-D and 2D training data
df_all <- dplyr::bind_cols(training_data |> dplyr::select(-ID), UMAP_data_with_hb_id)

## Averaged on high-D
df_bin <- avg_highD_data(.data = df_all)

## Triangulate bin centroids
tr1_object <- triangulate_bin_centroids(df_bin_centroids, x, y)
tr_from_to_df <- generate_edge_info(triangular_object = tr1_object)

## Compute 2D distances
distance <- cal_2D_dist(.data = tr_from_to_df)

## To find the benchmark value
benchmark <- find_benchmark_value(.data = distance, distance_col = distance)

trimesh <- ggplot(df_bin_centroids, aes(x = x, y = y)) + 
    geom_segment(data = tr_from_to_df, aes(x = x_from, y = y_from, xend = x_to, yend = y_to)) +
    geom_point(size = 2, colour = "#33a02c") +
    coord_equal()

  # ggplot(df_bin_centroids, aes(x = x, y = y)) + 
  # geom_point(size = 1, colour = "#33a02c") + 
  # geom_trimesh() + 
  # coord_equal()

trimesh <- trimesh +
  #ggtitle("(a)") + 
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme_light() +
    theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
    ) 
  # theme(axis.text = element_text(size = 5),
  #       axis.title = element_text(size = 7))

trimesh_gr <- colour_long_edges(.data = distance, benchmark_value = benchmark, 
                                triangular_object = tr1_object, distance_col = distance)

trimesh_gr <- trimesh_gr + 
    geom_point(size = 2, colour = "#33a02c") + 
  #ggtitle("(b)") + 
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme_light() +
    #coord_equal() +
    theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
    )

trimesh_removed <- remove_long_edges(.data = distance, benchmark_value = benchmark, 
                                triangular_object = tr1_object, distance_col = distance)

trimesh_removed <- trimesh_removed + 
       geom_point(size = 2, colour = "#33a02c") + 
  #ggtitle("(b)") + 
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme_light() +
    #coord_equal() +
    theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
    )

tour1 <- show_langevitour(df_all, df_bin, df_bin_centroids, benchmark_value = benchmark, distance = distance, distance_col = distance)
```

```{r}
#| echo: false

## To plot the distribution of distance
plot_dist <- function(distance_df){
  distance_df$group <- "1"
  dist_plot <- ggplot(distance_df, aes(x = group, y = distance)) +
    geom_quasirandom()+
    ylim(0, max(unlist(distance_df$distance))+ 0.5) + coord_flip()
  return(dist_plot)
}
```

```{r}
#| echo: false

distance_plot <- plot_dist(distance) +
  #ggtitle("(b)" ) + 
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))
```

## Notations

| Notation                                                                                                              | Description                                                            |
|-----------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------|
| $n$                                                                                                                   | number of observations                                                 |
| $p$                                                                                                                   | number of dimensions in high-D data                                    |
| $X_{n \times p} = \begin{bmatrix} \textbf{x} _{1} & \textbf{x}_ {2} & \cdots & \textbf{x}_{n} \\  \end{bmatrix}^\top$ | high-D data matrix                                                     |
| $d$                                                                                                                   | number of dimensions in embedding data, usually two                    |
| $Y_{n \times d} = \begin{bmatrix} \textbf{y} _{1} & \textbf{y}_ {2} & \cdots & \textbf{y}_{n} \\  \end{bmatrix}^\top$ | embedding matrix                                                       |
| $P_t$                                                                                                                 | point along the geodesic path at time $t$                              |
| $P_{t+1}$                                                                                                             | point along the geodesic path at the next time instant, which is $t+1$ |
| $b_1$                                                                                                                 | number of bins along the x-axis                                        |
| $b_2$                                                                                                                 | number of bins along the y-axis                                        |
| $b$                                                                                                                   | total number of bins in the hexagonal grid                             |
| $b'$                                                                                                                  | total number of bins without the empty bins                            |
| $h$                                                                                                                   | diameter of a hexagon                                                  |
| $A$                                                                                                                   | area of a hexagon                                                      |
| $r_1$                                                                                                                 | 2D embedding component 1                                               |
| $r_2$                                                                                                                 | 2D embedding component 2                                               |
| $s$                                                                                                                   | shape parameter                                                        |
| $C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$                                                                               | hexagonal bin centroid coordinates                                     |
| $C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})$                                                                          | averaged high-D coordinates                                            |
: Notations used in this paper and their descriptions. {#tbl-notations}

## Default parameter values

### Benchmark value to remove low-density hexagons

Removing all the bins that have a standard number of points less than a certain benchmark value will generate unnecessary holes within the triangular mesh. Therefore, before removing the low-density bins, it's necessary to investigate the number of points in neighboring bins as well.

$$
\text{standard count} = \frac{\text{count}}{\text{max count}} 
$${#eq-equationp2}

The process begins by identifying the six nearest hex bins of specified low-density hex bins. To identify the nearest hex bins 2D Euclidean distance metric is used. Then, we calculate the mean density, as outlined in the equations:

$$
\text{mean density} = \frac{\text{standard count}}{6} 
$${#eq-equationp3}

The standard count is derived from the number of observations in the hex bins. By examining the distribution of mean densities and designating the first quartile as the benchmark value, hex bins with mean densities below this benchmark are removed. This process ensures the elimination of regions with insufficient data density, focusing on areas with more significant data representation and preserving the overall structure in the low-dimensional space.

### Benchmark value to remove long edges {#sec-rmlgdefault}

Determining a benchmark value for removing long edges isn't a straightforward process with standard guidelines. Having a default value provides a practical starting point for decision-making, helping users choose whether to stick with the default, choose a value below it, or select one above it for long edge removal.

To calculate the default value, the process involves computing the 2D Euclidean distances between the vertices in the triangular mesh. Initially, distances between vertices are sorted. Then, unique distance values are extracted. Following this, the differences between consecutive distance values are computed, and a data frame is constructed with unique distance values and their corresponding differences. The first largest difference is then used as a benchmark to identify and remove the long edges.

![A flow diagram detailing the steps taken to find the benchmark value to remove long edges.](figures/remove_long_edges_workflow.png){#fig-rmlgmeth fig-align="center" width="100%" height="100%"}

```{r}
#| echo: false

dist_p <- plot_dist(distance)  +
  geom_hline(yintercept = benchmark, colour = "blue") + 
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12)) 
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Is the default benchmark value effective in removing the long edges to construct a smooth surface in 2D? (a) Triangular mesh with coloured edges by their type, (b) Distribution of 2D Euclidean distances between triangular mesh vertices."
#| label: fig-defaultrmlg
##| out-width: 100%

trimesh_gr + dist_p +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

## Paramter tunning

### Total number of bins

The total number of bins represents the overall count of hexagonal bins within the hexagonal grid. This count is determined by multiplying the number of bins along the x-axis ($b_1$) with the number of bins along the y-axis ($b_2$), according to the formula:

$$
 b = b_1 \times b_2
$$ {#eq-equationp1}

Here, $b$ denotes the total number of bins. By adjusting the parameter $b_1$, we have control over the total number of bins ($b$). Hence, fine-tuning $b_1$ enables us to customize and optimize the total bin count based on the desired configuration along the x-axis.

```{r}
#| echo: false
#| message: false
#| warning: false

hexgrid_with_points_hex <- ggplot(data = hex_full_count_df |> dplyr::filter(!is.na(std_counts)), aes(x = c_x, y = c_y, fill = std_counts, hexID = hexID)) + 
    geom_hex(stat = "identity", color = "#969696") +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  theme_light() +
    #coord_equal() +
    theme(legend.position = "none", plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
    )

@@

```{r}
#| echo: false
#| message: false
#| warning: false

hexgrid_with_points_m <- ggplot(data = hex_full_count_df, aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  theme_light() +
    #coord_equal() +
    theme(plot.title = element_text(size = 5, hjust = 0.5, vjust = -0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank()#change legend key width
    )

hexgrid_with_points <- hexgrid_with_points_m + theme(legend.position = "none")
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| out-width: 100%

hexgrid_with_points_hex + hexgrid_with_points +
  plot_layout(ncol = 3)

```
<!--
```{r}
#| echo: false
#| message: false
#| warning: false
 # Loop over loop.vector
  #i <- 1
  num_bins <- 3
  shape_val <- calculate_effective_shape_value(.data = UMAP_data, x = UMAP1, y = UMAP2)

  hexbin_data_object_loop <- extract_hexbin_centroids(UMAP_data, num_bins, shape_val)

  df_bin_centroids_loop <- hexbin_data_object_loop$hexdf_data

  ## Data set with all possible centroids in the hexagonal grid

  full_centroid_df_loop <- generate_full_grid_centroids(df_bin_centroids_loop)

  ## To map hexID to hexbin centroids in the full grid

  vec1 <- stats::setNames(rep("", 2), c("x", "y"))  ## Define column names

  full_grid_with_hexbin_id_loop <- dplyr::bind_rows(vec1)[0, ]
  full_grid_with_hexbin_id_loop <- full_grid_with_hexbin_id_loop |>
    dplyr::mutate_if(is.character, as.numeric)

  for(i in 1:length(sort(unique(full_centroid_df_loop$y)))){

    ## Filter the data set with specific y value
    specific_y_val_df_loop <- full_centroid_df_loop |>
      dplyr::filter(y == sort(unique(full_centroid_df_loop$y))[i])

    ordered_x_df_loop <- specific_y_val_df_loop |>
      dplyr::arrange(x)

    full_grid_with_hexbin_id_loop <- dplyr::bind_rows(full_grid_with_hexbin_id_loop, ordered_x_df_loop)

  full_grid_with_hexbin_id_loop <- full_grid_with_hexbin_id_loop |>
    dplyr::mutate(hexID = row_number())

  full_grid_with_hexbin_id_loop <- full_grid_with_hexbin_id_loop |>
    dplyr::rename("c_x" = "x",
           "c_y" = "y")

  full_grid_with_hexbin_id_loop <- dplyr::full_join(full_grid_with_hexbin_id_loop, df_bin_centroids_loop, by = c("hexID" = "hexID")) |>
    dplyr::select(-c(x, y))

  full_grid_with_hexbin_id_loop <- full_grid_with_hexbin_id_loop |>
      dplyr::mutate(std_counts = counts/max(counts, na.rm = TRUE))

  ## Generate all coordinates of hexagons
  hex_grid_loop <- full_hex_grid(full_centroid_df_loop)

  full_grid_with_polygon_id_df_loop <- map_polygon_id(full_grid_with_hexbin_id_loop, hex_grid_loop)

  full_grid_with_hexbin_id_rep_loop <- full_grid_with_polygon_id_df_loop |>
  dplyr::slice(rep(1:n(), each = 6)) |>
  dplyr::arrange(polygon_id)

  hex_full_count_df_loop <- dplyr::bind_cols(hex_grid_loop, full_grid_with_hexbin_id_rep_loop)

  p1 <-  ggplot(data = hex_full_count_df_loop, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
        geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
    theme_void() +
    theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
          axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
          legend.title = element_text(size=8), #change legend title font size
          legend.text = element_text(size=6)) +
    guides(fill = guide_colourbar(title = "Standardized count")) +
    annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

>>>>

```{r}
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)
library(tidyr)

library(Rtsne)
library(uwot)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
set.seed(20240110)

source("nldr_code.R", local = TRUE)
```

<!-- To find the number of bins should assign for the model in tSNE-->
```{r}
#| warning: false
#| echo: false
#| message: false

training_data_gau <- read_rds("data/five_gau_clusters/data_five_gau_training.rds")
tSNE_data_gau <- read_rds("data/five_gau_clusters/tsne_data_five_gau_61.rds")

tsne_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tSNE_data_gau, 
                                    x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tSNE_data_gau))

## tSNE
hex_size_vec <- seq(0.02, 2, by = 0.01)

vec <- stats::setNames(rep("", 6), c("num_bins", "aic", "mse", "num_bins_x", "num_bins_y", "hex_size"))  ## Define column names

mse_df_gau <- dplyr::bind_rows(vec)[0, ]
mse_df_gau <- mse_df_gau |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(hex_size_vec)) {

  num_bin_list <- calc_bins(data = tsne_gau_scaled, 
            x = "tSNE1", y = "tSNE2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  num_bins_x <- num_bin_list$num_x
  num_bins_y <- num_bin_list$num_y

  model_object <- fit_highd_model( training_data = training_data_gau, 
                                   nldr_df_with_id = tsne_gau_scaled, 
                                   x = "tSNE1", y = "tSNE2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "tSNE", 
                                   col_start_highd = "x")

  pred_emb_list <- predict_emb(test_data = training_data_gau, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list))

  eval_list <- gen_summary(test_data = training_data_gau, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "x")

  mse_df_gau <- mse_df_gau |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    aic = eval_list$aic,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i])

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df_gau |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df_gau |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain one row from duplicates which have lowest error and hexsize
duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))

for (i in 1:length(duplicate_df_list)) {

  dd <- duplicate_df_list[[i]] |>
    dplyr::filter(mse == min(duplicate_df_list[[i]]$mse)) |>
    dplyr::filter(hex_size == min(duplicate_df_list[[i]]$hex_size))

  duplicate_df <- dplyr::bind_rows(duplicate_df, dd)

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_gau |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_gau <- dplyr::bind_rows(not_dupli_df, duplicate_df) 
```

@@

mse_plot_gau <- ggplot(mse_df_gau, aes(x = num_bins,
                                       y = log(mse)
)) +
  geom_point(size = 0.8) +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from different NLDR techniques applied to training five spherical Gaussian cluster dataset. What is the best NLDR technique to represent the original data in 2D?"
#| label: fig-diagnosticpltGau
#| out-width: 100%
#| fig-pos: H

mse_plot_gau 
```

<!-- MSE plot to find which is the effective number of bins to use-->
```{r}
#| warning: false
#| echo: false
#| message: false

## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                                                 x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

#tot_num_bin_vec <- 2:51
hex_size_vec <- seq(0.02, 2, by = 0.01)

vec <- stats::setNames(rep("", 7), c("num_bins", "error", "mse", "num_bins_x", "num_bins_y", "hex_size", "num_non_empty_bins"))  ## Define column names

mse_df_pbmc <- dplyr::bind_rows(vec)[0, ]
mse_df_pbmc <- mse_df_pbmc |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(hex_size_vec)) {

  num_bin_list <- calc_bins(data = umap_pbmc_scaled, 
            x = "UMAP1", y = "UMAP2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  num_bins_x <- num_bin_list$num_x
  num_bins_y <- num_bin_list$num_y

  model_object <- fit_highd_model( training_data = training_data_pbmc, 
                                   nldr_df_with_id = umap_pbmc_scaled, 
                                   x = "UMAP1", y = "UMAP2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "UMAP", 
                                   col_start_highd = "PC_")

  pred_emb_list <- predict_emb(test_data = training_data_pbmc, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "UMAP")

  pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list))

  eval_list <- gen_summary(test_data = training_data_pbmc, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "PC_")

  mse_df_pbmc <- mse_df_pbmc |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    error = eval_list$error,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i],
                    num_non_empty_bins = NROW(centroid_df_training))

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df_pbmc |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df_pbmc |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain one row from duplicates which have lowest error and hexsize
duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))

for (i in 1:length(duplicate_df_list)) {

  dd <- duplicate_df_list[[i]] |>
    dplyr::filter(mse == min(duplicate_df_list[[i]]$mse)) |>
    dplyr::filter(hex_size == min(duplicate_df_list[[i]]$hex_size))

  duplicate_df <- dplyr::bind_rows(duplicate_df, dd)

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_pbmc |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_pbmc <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

@@

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training PBMC3k dataset. What is the effective number of bins in UMAP to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 588, the slope of the MSE plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-msepltpbmc
#| fig-pos: H

mse_plot_pbmc
```

<!-- MSE plot to find which is the effective number of bins to use-->

```{r}
#| warning: false
#| echo: false
#| message: false

## Import data
training_data_mnist <- read_rds("data/mnist/mnist_10_pcs_of_digit_1.rds")
training_data_mnist <- training_data_mnist |>
  mutate(ID = 1:NROW(training_data_mnist))

pacmap_minst <- read_rds("data/mnist/mnist_pacmap.rds")
pacmap_minst_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_minst, 
                                    x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1", 
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_minst))

#tot_num_bin_vec <- 2:51
hex_size_vec <- seq(0.02, 2, by = 0.01)

vec <- stats::setNames(rep("", 6), c("num_bins", "aic", "mse", "num_bins_x", "num_bins_y", "hex_size"))  ## Define column names

mse_df_mnist <- dplyr::bind_rows(vec)[0, ]
mse_df_mnist <- mse_df_mnist |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(hex_size_vec)) {

  num_bin_list <- calc_bins(data = pacmap_minst_scaled, 
            x = "PaCMAP1", y = "PaCMAP2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  num_bins_x <- num_bin_list$num_x
  num_bins_y <- num_bin_list$num_y

  model_object <- fit_highd_model( training_data = training_data_mnist, 
                                   nldr_df_with_id = pacmap_minst_scaled, 
                                   x = "PaCMAP1", y = "PaCMAP2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "PaCMAP", 
                                   col_start_highd = "PC")

  pred_emb_list <- predict_emb(test_data = training_data_mnist, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "PaCMAP")

  pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list))

  eval_list <- gen_summary(test_data = training_data_mnist, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "PC")

  mse_df_mnist <- mse_df_mnist |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    aic = eval_list$aic,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i])

@@

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df_mnist |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df_mnist |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain one row from duplicates which have lowest error and hexsize
duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))

for (i in 1:length(duplicate_df_list)) {

  dd <- duplicate_df_list[[i]] |>
    dplyr::filter(mse == min(duplicate_df_list[[i]]$mse)) |>
    dplyr::filter(hex_size == min(duplicate_df_list[[i]]$hex_size))

  duplicate_df <- dplyr::bind_rows(duplicate_df, dd)

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_mnist |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_mnist <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

mse_plot_mnist <- ggplot(mse_df_mnist, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 180, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from PaCMAP applied to training MNIST dataset. What is the effective number of bins in PaCMAP to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 180, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltMNIST
#| fig-pos: H

mse_plot_mnist
```
